---
title: 'Designometric models for the evaluation of designs.'
author: "Martin Schmettow and Simone Borsci"
output:
  html_document:
    number_sections: yes
---

```{r setup, warning = FALSE, message = FALSE, echo = T}
library(tidyverse)
library(psych)
library(mascutils)
library(printr)
#library(lavaan)
options(mc.cores = 5)

purp.analysis <- T


# D_1 |> 
#   filter(Scale == "Attractiveness") |> 
#   item_rel()


## Functions to create response matrices from long data

rm_psycho <- function(Data)
  Data |> 
  group_by(Part, Item) |> 
  summarize(mean_resp = mean(response)) |> 
  ungroup() |> 
  arrange(Item) |> 
  spread(Item, value = mean_resp) |> 
  select(-Part)

rm_design <- function(Data) 
  Data |> 
  group_by(Design, Item) |>
  summarize(mean_resp = mean(response)) |> 
  ungroup() |> 
  spread(Item, value = mean_resp) |> 
  select(-Design)


alpha_ci <- function(Data){
  Scale <- str_c(distinct(Data, Scale)$Scale) 
  model_psych <- 
    psych::alpha(rm_psycho(Data), check.keys = FALSE, n.iter = 100)$boot |> 
    as_tibble() |> 
    mutate(Perspective = "psychometric")
  model_design <- 
    psych::alpha(rm_design(Data), check.keys = FALSE, n.iter = 100)$boot |> 
    as_tibble() |> 
    mutate(Perspective = "designometric")
  out <- 
    bind_rows(model_psych,
              model_design) |> 
    select(Perspective, std.alpha) |> 
    group_by(Perspective) |> 
    summarize(center = mean(std.alpha),
              lower = quantile(std.alpha, .025),
              upper = quantile(std.alpha, .975)) |> 
    mutate(Scale = Scale) |> 
    mascutils::go_first(Scale, Perspective)
  out
}

# D_1 |> 
#   filter(Scale == "Attractiveness") |> 
#   alpha_ci()


item_rel <- function(Data){
  #Data <- D_1 |> filter(Scale == "HQI")
  Scale <- str_c(distinct(Data, Scale)$Scale)
  model_psych <- 
    psych::alpha(rm_psycho(Data), check.keys = FALSE)$item.stats |> 
    as_tibble(rownames = "Item") |> 
    mutate(Perspective = "psychometric")
  model_design <- 
    psych::alpha(rm_design(Data), check.keys = FALSE)$item.stats |> 
    as_tibble(rownames = "Item") |> 
    mutate(Perspective = "designometric")
  
#  model <- M_1_design
  out <- 
    bind_rows(model_psych,
              model_design) |> 
    mutate(Scale = Scale) |> 
    go_first(Scale, Item, Perspective) |> 
    arrange(Scale, Item, Perspective)
  out
}




```

# Introduction


<!-- It was a wonder to observe how the pale definition of user satisfaction had to make space for a big party. And everybody was bringing their own rating scales! (Bargas-Avila & Hoernbaek, Old wine in new bottles) -->

In modern industrial practice, rating scales have their place as an always available and cheap method for comparing or benchmarking designs. In the decision process, the everyday value of a rating scales stands and falls with two properties: validity and reliability.

*Developing* a valid and reliable rating scale is quite an undertaking. *Psychometrics* is the science of assigning numbers to persons so that they can be compared by psychological functioning. Traditionally, this served to measure skills, such as mathematical intelligence or comprehension of language. With the time, researchers also became more interested in elusive properties of persons, such as psycho-social tendencies (known as the Big Five).

After the landrush phase of the UX revolution, Bargas-Avila & Hoernbaek \[Old wine in new bottles\] counted hundreds of new rating scale instruments. Barely any of these instrument underwent the same scrutiny as, for example, a rating scale for psycho-pathological diagnosis would have to. But, frequently some psychometric tools were used at one point during development time, for example by reporting reliability estimates. Also for users of such instruments, it is common to perform basic psychometric sanity checks on their data. The central point of this paper is a certain catch that can occur when translating between psychological research and design research, which we call the *psychometric fallacy*.

In psychometric situations, the atomic observation is an encounter of a person with a test item. This is repeated with more items to improve the precision of the measurement. If many persons are assessed this way, the result is a Person X Item *response matrix*, from which *person scores* can be extracted for ranking. *Item scores*, are mostly used during scale development. 

The logical argument developed in this paper is that instruments in design research exist to rank designs and the atomic observation is an encounter of a design with a person and an item. This forms a box of data, which can be collapsed by three perspectives. We demonstrate how standard psychometric tools can be used correctly using the designometric response matrix. We call it the psychometric fallacy, when a psychometric matrix is used instead.

...

Aim of this study is also to seek empirical evidence that the psychometric fallacy is not just sophistry, but can result in real biases when developing or using rating scales. For this purpose, data from five experiments was subjected to typical rating scale validation techniques under both perspectives, psychometric (pretending the fallacy) and designometric (using the proper response matrix).

## Psychometrics

It only takes a single probe to accurately estimate the body temperature of a person. Why are psychological properties of the same person, e.g. language skills or neuroticism, assessed by multiple tasks or items? While there is more than one reason for multiple items in questionnaire, but the one primarily addressed here is the measurement accuracy. Data taken from the human mind is highly noisy, be it obtained by self-report scales, reaction times or physiological measures. Practically all quantitative behavioural studies employ repeated measures to reduce the measurement error.

This goes back to so-called *classic test theory*, which states that in behavioural research the level of noise in a single measurement usually is too large all items contain an amount of error, which is usually to large to be practical. Classic test theory solves this problem by decomposing every measure into a relevant *systematic component* and a nuisance *error component*. By definition, the systematic component re-occurs with every measurement, whereas the error is fully random. With a single item it is impossible to separate these components, but the more items are used, the more the error terms cancel each other out, and the systematic component. When certain conditions are met, it is possible to create an arbitrary precise estimate by repeating the measurement.

The primary use of psychometric instruments is to make *predictions* about a person, which in turn helps to make the best decision about that person. A good score in a written drivers test predicts that a person will make fewer errors on the road and experience fewer collisions.

<!--A more direct way to measure error-proneness would be to send the person into traffic and count the collisions over a long period of time. As illustrated by this extreme example, a good instrument also needs to be cost-effective.-->

Psychometric instruments can be seen as simulations of real situations that supposedly trigger the same areas of psychological functioning, but in a safe and cost-efficient way.

### Item selection

The first challenge when designing a psychometric inventory is to understand the domain, ideally in terms of involved areas of psychological functioning. Based on the domain analysis, the researcher creates a candidate *item pool*, which is usually much larger than the targeted final set of items. Several psychometric methods can be used to successively remove inefficient items, and to find an effective structure.

In the ideal case, the errors of multiple items cancel each other out. This requires that all items share the same systematic component. Scale *consistency* is the overall level of inter-item agreement, where lack of consistency occurs when a single item fails to correlate with other items, simply because it is ill-phrased or ambivalent. Inconsistent individual items can be identified using item reliability measures, and removed from the pool.

Reliability of an item is the essentially the inverse of the error component and denotes how precisely a measurement predicts itself. As the error component is unknown, reliability can only be estimated. For a single item test, the only way is to correlate a test with a retest. But, multi-item scales have the advantage that reliability can be estimated in one moment, with Cronbach $\alpha$ as a common technique. Cronbach $\alpha$ correlates each item with the remaining items, which results in reliability measures on item level.

In more complex cases, scale consistency suffers, because there is more than one systematic component. In this case, *factor-analytic* methods can be used to separate components into consistent factors.

### Factor structures

In modern psychometrics, the relation between a supposed true property and multiple imperfect measure is known as the distinction between *latent variables* (not observable, but true) and *indicator variables* (observable, but imperfect). Modern instruments often carry a domain structure, where multiple latent variables are put in relation to each other. A prominent example is the Big 5 inventory, which claims that social behaviour can be predicted by five psycho-social traits, each of which is assessed by a separate set of items.

While the primary aim of domain analysis is to gain sufficient coverage of the domain, it often at least indicates a possible structure for a multi-item instrument. In some cases, a domain analysis is driven by a theoretical structure in the first place. For example, a rating scale for mental workload could very well be based on the multiple-resource theory, which predicts that different sensory modes are processed independently. This simply translates to one scale per sensory mode.

1.  Auditory load: *It is difficult to understand the signal tones.*
2.  Visual load: *The amount of visual information can be overwhelming*.

It is even possible to derive a total score for this situation, which is not the usual mean, but *the maximum*. Most

Prior structures can also emerge from qualitative data in mixed-method psychometric processes. For example, Confirmatory Factor Analysis (CFA) is the most recommended technique to critically assess *a prior* assumptions about the latent variable structure. For multi-scale inventories CFA also supercedes Cronbach $\alpha$, as parameter loadings can be used to identify inconsistent items.

Without a prior structures, the single-scale process of item selection is likely to fail the consistency test, because the alleged skill is actually a composition of skills. For example with open questions in an arithmetics test, performance may depend on arithmetic processing speed, but also on comprehension. If items in the test vary in text difficulty (and participants vary in comprehension skill), a second systematic component emerges. A common technique for finding new structures is Exploratory Factor Analysis. An issue with EFA is that it requires the researcher to name the number of factors. Several methods have been proposed to identify the optimal number of factors before-hand.

### The Response Matrix

All psychometric methods mentioned so far have in common that measures are provided in form of response matrices, where rows identify with participants and columns with items. A common operation during item selection is to estimate consistency on the full response matrix, remove a dubious item from the matrix to see if this improves consistency. EFA takes a response matrix as input and produces item loadings for it. For multi-scale data CFA tools take response matrix with named columns as input, together with a formula to group items to factors. ...

The idea of response matrices is most pronounced in item response theory (IRT), a psychometric approach rivaling classic test theory and factors analysis. In IRT, each cell in the response matrix is seen as an encounter between an item and a participant, from which a measure arises. Items, like participants, are seen as individuals, with different response characteristics during the encounter. In the basic Rasch model, the probability of success in a person-task encounter depends only on the general difficulty of task and the general ability of the person. More complex IRT models also regard more subtle item characteristics, and sophisticated tools have been developed to detect and prevent *differential item functioning* in world-spanning studies like TIMMS and PISA. IRT models take the person x item matrix as input and produce estimates on both margins, item parameters and person parameters. During the development of an IRT model, sophisticated tools are used to select only those items that strictly behave as the model states.

For all levels of sophistication or schools of thought, what counts during scale development is that item characteristics are proven to be well-behaved according to theory. In the same way teachers use multiple tests to estimate a student's skill, many responses are needed to precisely assess the characteristics of an item during this phase. This creates the demand for the enormous sample sizes, that psychometric development is most feared for.

### Sample sizes

The enormous sample sizes consumed during psychometric scale validation have several causes:

-   Psychometric techniques tend to have many parameters and by the rule that the number of independent observations must at least match the number of parameters, a lower boundary can be established. A 24 item three-way EFA already produces 72 parameters. But, during scale validation, the number of items is often three times larger, which adds up.
-   The estimation also has to deal with strong noise, which creates the practical problem that certain estimation procedures need a lot of data to converge properly, for example in consitional maximum likelihood estimation.
-   When a structure is desired, but no prior structure exists, EFA can be used to establish a structure. But the data used for EFA is consumed. To confirm the structure, a new sample is required.
-   The main areas of psychometric research are education and clinical diagnosis, where a rating scale or test can have a significant impact on a person's life. These instruments ought to be extra-hardened against biases and require the greatest scrunity. For item in a PISA test, proof is required that it has the same characteristics all around the globe.

## Designometrics

Everything said above is essentially rooted in mathematical theory and therefore equally holds when rating scales are constructed to rank humans, but also designs, and anything else. When a UX scale is used to decide between designs A and B in industrial practice, it is legit to ask how precise the measurements are. Simple, low-stake Dmx scales can efficiently be constructed and evaluated using Cronbach $\alpha$. For multi-scale inventories, factor analysis can be used. And when maximum rigor is required, the scale can be tested against an IRT model.

There are just two differences: First, in psychometrics the entity to be ranked is persons, whereas designometric applications one or more designs get numbers attached. Second, a psychometric measure is an encounter of a person with an item, whereas a designometric measure is an encounter of a design with a person and an item. The three-way encounter and designs being subject to measurement is what defines *designometric situations*.

### The designometric perspective

A complete designometric data set is a cross-product of three samples, which would require a box-like input form, rather than a matrix, and can not directly be processed with psychometric tools.

By simply averaging across one factor, box data can be transformed into a matrix to fit into psychometric tools. For designometric instrument validation, averaging across Person creates a design-by-item *designometric response matrix*. This matrix takes the place of the psychometric response matrix and researchers can use it with standard psychometric tools to select proper items and find good structures.

At the same time, a designometric box can also be seen as a *stack of psychometric response matrices*, where averaging across the Design factors produce a psychometric response matrix.

### Designometric sampling

When Design takes the place of Person, it is implied that items must now be well-behaved in ranking designs, and substantial samples of designs are now required to prove that. This can be a minor or huge problem, depending on the class of designs. For e-government websites a large design sample will be easier to come by than a sample of human-like robots or self-driving cars.

A connected problem is that the designometric perspective participant-level information is reduced, but they are still involved in the designometric encounter. In the special case that scales are used for first impression ratings, it is possible to rund cross-product complete designometric encounters. For example, when we used the Eeriness scale on artificial faces, a single observation only took around 4 seconds, making it relatively easy to fill the designometric box with data, even repeatedly \[BT Robbin Koopmans\].

When a real interactive experience is subject of the measure, a measurent can take from several minutes to hours and a complete designometric encounter becomes impractical. A way to mitigate this problem is to use an experimental design that is *planned incomplete*. Essentially, a planned incomplete validation study has all participants encounter only a partition of the design sample. For example a study consisting of a sample of 100 designs let every participant encounter a different set of ten designs. As long as all designs are covered by at least one participant, this will result in a complete design-by-item matrix after collapsing along participants.

A variation of planned incomplete studies is to *successively* build the sample of designs. This is especially useful, when dealing with emerging classes of designs. This happened in the BUS-11 studies, where initially it was difficult to build a substantial sample, before it became more common.

## The psychometric fallacy

Designometric scales can be developed and validated using established psychometric tools. However, this is only valid when a design-by-item matrix is used, with a sufficiently large sample of designs. In contrast, many designometric instruments have not been validated using a large sample of designs, but rather on a large psychometric matrix. This we call the *psychometric fallacy*.

When a scale validation study in design research is in the psychometric fallacy, validation metrics such as item reliability may be meaningless for the purpose of ranking designs. Rather, the metric will refer to the capability of the item to discriminate persons by their sensitivity to the design feature in question. For example, a scale for comparing designs by beauty would become a scale to rank persons by how critical they are with respect to interface beauty. This is not the same and in the next section we show by simulation that the differences between designometric and psychometric perspectives can be dramatic.

During the scale development process, the psychometric fallacy appears in two forms, one is repairable, whereas the other is fatal. We speak of a repairable psychometric fallacy, when the validation study properly collected designometric data, but used psychometric response matrices for validation. This can be repaired by redoing the analysis now using a designometric matrix.

A study that did not collect a sample of designs, but used only a single or very few designs fell for the *fatal psychometric fallacy*. In these cases, researchers have failed to recognize a simple truth: The capacity to discriminate between designs can impossibly be validated on a single design. Every alleged designometric instrument, where this has happened during the validation phase, cannot be trusted.

Recall that psychometric validations require large samples of participants! Similar sample sizes will be required for designs when validating designometric instruments. So, even validation studies that included multiple designs, may in practice not be repairable, because the sample of designs is too small for that type of analysis (e.g. factor analysis). To give an example \[MacDorman XY\] validated the Godspeed Index, a common inventory to evaluate robotic designs. While scale reliability was assessed under a psychometric perspective, validity was correctly assessed using a sample of four designs. This allowed the authors to do a rough test on the ranking capabilities of their scales, but would not be suitable assess more detailed metric properties of the scale.

As a milder form *run-time psychometric fallacy* appears when an instrument is used in practice to take measures on a single design. The result will inevitable look like a psychometric response matrix and, given that publication rules (e.g. APA guidelines) often require to report some psychometric properties, it may be tempting for the researcher to run a psychometric test on reliability. While the run-time fallacy does not have the same impact as development-time fallacies, it may have cause some head aches when a validated instrument seems to have poor reliability.

In the following section, we construct a case by simulation, showing that psychometric and designometric perspectives can result in dramatically different results. In the remainder of the study, we will use data from past experiments to evaluate how strong the deviations are with several existing and commonly used rating scales.

# Simulation study

The following example demonstrates how an extreme situation can arise, when a fictional four-item scale for Coolness is highly reliable for persons, but has no reliability at all for ranking designs. By an unfortunate choice, the sample of designs consisted were exclusively homepages of premium law firms, with little variation ($\sigma_\textrm{Design} = .02$) on a low level ($\mu_\textrm{Design} = -.5$).

In this simulation, we assume that the Coolness scale has been tested on a sample of 50 designs and 50 participants, where . The key here is that participants vary strongly in their sensitivity to Coolness ($\sigma_\textrm{Part} = .2$), whereas the sample of designs varies little in Coolness .

```{r}
set.seed(42)

n_Design = 20
n_Part   = 20
n_Item  =  4
n_Obs = n_Design * n_Part * n_Item

Designs <- tibble(Design      = as.factor(1:n_Design),
                  cool_Design = rnorm(n_Design, -0.5, .02)) ## low level, little variation

Parts   <- tibble(Part        = as.factor(1:n_Part),
                  cool_Part   = rnorm(n_Part, 0, .2)) ## strong variance in tendency to judge sth. cool

Items   <- tibble(Scale       = "Coolness",
                  Item        = as.factor(1:4),
                  cool_Item   = rnorm(n_Item,  0, .2)) ## item strength: understating items get lower values

Coolness     <- expand_grid(Design = Designs$Design,
                       Part   = Parts$Part,
                       Item   = Items$Item) |> 
  left_join(Designs) |> 
  left_join(Parts) |> 
  left_join(Items) |> 
  mutate(response = mascutils::rescale_zero_one(cool_Design + cool_Part - cool_Item + rnorm(n_Obs, 0, .5)))
```

```{r eval = F, echo = F}
Coolness |> 
  ggplot(aes(y = response, x = Design)) +
  geom_violin()

Coolness |> 
  ggplot(aes(y = response, x = Part)) +
  geom_violin()
```

```{r}
alpha_ci(Coolness)
```

This simple example demonstrate that a scale can produce excellent reliability when measuring person sensitivity, but poor and uncertain reliability on designs. Under the psychometric fallacy it can happen that excellent reliability is reported, while it is actually unknown, or very poor.

In the following study we use data from several previous experiments that had produced designometric data sets.

# Methods

From a theoretical perspective the psychometric fallacy is obvious and we have demonstrated by simulation that the worst case is possible , but little is known how the fallacy effects the quality of rating scales. Here, we explore the biases that can occur when a psychometric response matrix is used, in place of a designometric. By collecting from previous experiments, a sample of UX scales was put to the test with what constitutes a basic set of tools for devloping multi-scale instrument.

## Data sets

The data used for analysis originates from five experiments (DK, PS, AH, QB, DN). While these experiments were carried out to test their own hypotheses, they have in common that participants saw pictures of many designs and were asked to respond to items taken from one or more scales. In QB and DN participants saw pictures of home pages and responded to several user experience scales, whereas in AH, DK and PS the stimuli were robot faces. Some of the original experiments used manipualtion of presentation time to collect data on subconscious cognitive processing. For the analysis here, only responses at presentation times of 500ms and 2000m were used.

Per trial participants saw a single design followed by a random single item, resulting in a sparse designometric box. However, when collapsing the box to either psychometric RM or designometric RM, the result is completely filled response matrices.

```{r}
load("DMX_data.Rda")
```

```{r}
D_1 |> 
  ggplot(aes(x = response)) +
  geom_histogram() +
  facet_wrap(~Scale, scale = "free_y")
```

```{r eval = F, echo = F}
D_Eer |> 
  ggplot(aes(x = Item, y = response, color = Study)) +
  geom_violin()
```

```{r echo = purp.analysis}
D_1 |> 
  distinct(Study, Design) |> 
  group_by(Study) |> 
  summarize(n_Design = n()) |> 
  ungroup() |> 
  left_join(D_1 |> 
              distinct(Study, Part) |> 
              group_by(Study) |> 
              summarize(n_Part = n())|> 
              ungroup()) |> 
  left_join(D_1 |> 
              group_by(Study) |> 
              summarize(n_Obs = n())|> 
              ungroup()
  )
```

```{r echo = purp.analysis}
D_1 |> 
  distinct(Scale, Design) |> 
  group_by(Scale) |> 
  summarize(n_Design = n()) |> 
  ungroup() |> 
  left_join(D_1 |> 
              distinct(Scale, Part) |> 
              group_by(Scale) |> 
              summarize(n_Part = n())|> 
              ungroup()) |> 
  left_join(D_1 |> 
              group_by(Scale) |> 
              summarize(n_Obs = n())|> 
              ungroup()
  )
```

### Scales

For the following rating scales responses have been extracted from the original experimental data:

The *Eeriness* scale has been developed for measuring negative emotional responses towards robot faces and is a primary research tool on the Uncanny Valley phenomenon. Ho & MacDorman(2017) present an advanced psychometric validation of the scale. The study made use of 12 animated characters (Designs), avoiding the fatal fallacy to some degree, but the data analysis is under psychometric perspective.

The *Attractiveness* scale is part of the User Experience Questionnaire (UEQ) inventory. Is has been vaidated by \[Bettina Laugwitz, Theo Held, and Martin Schrepp. 2008. Construction and Evaluation of a User Experience Questionnaire. . 63–76. <https://doi.org/10.1007/978-3-540-89350-9_6>\] The UEQ has undergone basic psychometric evaluation in six studies with a single design each.

The two scales *Hedonic Quality - Identity (HQI)* and *Hedonic Quality - Stimulation (HQS)* are from the AttrakDiff2 inventory. AttrakDiff2 underwent basic evaluation using only three Designs under psychometric perspective (level 1 fallacy) \[Hassenzahl, M., Burmester, M., Koller, F., AttrakDiff: Ein Fragebogen zur Messung wahrgenommener hedonischer und pragmatischer Qualität\].

The Credibility scale ... \#### \[HERE\]


```{r echo = purp.analysis}
D_1 |> 
  group_by(Study, Scale) |> 
  summarize(n_Items = n_distinct(Item),
            n_Part = n_distinct(Part),
            n_Design = n_distinct(Design),
            n_Obs = n()) |> 
  ungroup()
```

### Data analysis

Goal of the analysis is to examine in how much the psychometric fallacy creates real biases. For this purpose, three basic psychometric techniques were applied to on several data sets: scale consistency

# Results

## Scale consistency

```{r message=FALSE, warning=FALSE}
Scale_rel <-
  D_1 |> 
  group_by(Scale) |> 
  group_split() |> 
  map_df(alpha_ci)

Scale_rel

```

```{r fig.height = 16, fig.width = 12}
Scale_rel |> 
  ggplot(aes(color = Scale,
             label = Scale,
             x = Perspective,
             y = center,
             ymin = lower,
             ymax = upper)) +
  geom_point() +
  geom_line(aes(group = Scale)) +
  ylab("std. Cronbach alpha") +
  geom_label() +
  ylim(0,1)
```

## Item reliability

```{r message=FALSE, warning=FALSE}
Item_rel <-
  D_1 |> 
  group_by(Scale) |> 
  group_split(Scale) |> 
  map_df(item_rel)

Item_rel

```

```{r eval = F, echo = F}
G_Item_rel <-
  Item_rel |> 
  ggplot(aes(color = Scale,
             x = Perspective,
             y = r.cor)) +
  # geom_point() +
  geom_line(aes(group = Item)) +
  ylab("Item-whole correlation")

G_Item_rel

```

```{r fig.width = 12, fig.height = 12}

Item_rel |> 
  ggplot(aes(x = Perspective,
             y = r.cor)) +
  geom_line(aes(group = Item)) +
  ylab("Item-whole correlation") +
  geom_label(aes( label = Item)) +
  facet_wrap(~Scale, ncol = 2) +
  geom_point(data = rename(Scale_rel, alpha = center),
             aes(x = Perspective, 
                 y = alpha,
                 col = "Whole Cronbach alpha")) +
  geom_line(data = rename(Scale_rel, alpha = center),
             aes(x = Perspective, 
                 y = alpha,
                 group = Scale,
                 col = "Whole Cronbach alpha"))


```

## Number of factors

Often, different scales are used in combination to create a more complete picture. It is usually the aim that a scale measures exactly one construct (or latent variable) and that different scales measure different constructs. As a counter-example, MacDorman found that the Eeriness scale decomposes into two slightly different aspects, summarized as "eery" and "spine-tingling". In contrast, the AttrakDiff2 questionnaire comprises two scales to capture supposedly different aspects.

Given a response matrix, the number of factors can be estimated using parallel analysis. Ideally, this procedure returns exactly as many factors as there are separate scales. Here, we use parallel analysis to assess whether the two perspectives produce the expected number of factors.

```{r}
parallel_analysis <- function(data, n, persp, scales){
  if (persp == "D") {
    data <- rm_design(data)
    main <- str_c("Designometric Parallel Analysis of ", scales)
  }
  if (persp == "P") {
    data <- rm_psycho(data)
    main <- str_c("Psychometric Parallel Analysis of ", scales)
  }
  psych::fa.parallel(data,
                   fa = "fa",
                   fm = "minres",
                   nfactors=n,
                   main=main)
    
}
```

<!-- Item reliability measures how coherent any particular item is with the scale as a whole. This criterion regularly used during item selection, where an item may be discarded if it does not correlate with the remaining items. However, if some items are out-of-tune, this can mean that they represent a different facet of the construct under examination.  -->

<!-- Exploratory factor analysis is frequently used to find such factors in response matrices. The first step typically is to determine the number of factors. -->

## Eeriness scale

Eeriness is usually considered a one-dimensional construct. Nevertheless, it has been suggested that it comprises two slightly different factors.

```{r}
parallel_analysis(D_Eer, 2, "D", "Eeriness")

parallel_analysis(D_Eer, 2, "P", "Eeriness")

```

The results suggest that under psychometric perspective there is only one latent variables, whereas whereas the designometric perspective produces two, indeed.

```{r}
efa <- function(rm, nfactors) psych::fa(rm_design(D_Eer),
                        fa = "fa", fm = "minres",
                        main=main)

Efa_psycho <- efa(rm_psycho(D_Eer), nfactors=1)
Efa_design <- efa(rm_design(D_Eer), nfactors=2)

psych::fa(rm_psycho(D_Eer))

E_psycho <- tibble(Perspective = "psychometric",
                   Item = str_c("nE", 1:8),
                   loading = psych::fa(rm_psycho(D_Eer))$loadings)

E_design <- tibble(Perspective = "designometric",
                   Item = str_c("nE", 1:8),
                   loading = Efa_design$loadings)

bind_rows(E_psycho, E_design) |> 
  ggplot(aes(x = Perspective, col = Item, y = loading)) +
  geom_point() +
  geom_line(aes(group = Item))
  



knitr::kable(bind_rows(E_psycho, E_design))
```



## AttrakDiff and Credibility

On theoretical grounds, the AttrakDiff2 inventory splits hedonistic quality into two components, Identity and Stimulation, while the credibility scale is a complete separate construct . Depending on whether Hedonism splits up like expected by theory, we would expect a two or three-factor solution.

```{r}
parallel_analysis(D_Att, 3, "P", "AttrakDiff and Credibility")

parallel_analysis(D_Att, 3, "D", "AttrakDiff and Credibility")

```

Under a psychometric perspective, all items represent a single latent construct. In contrast, the designometric analysis yielded five factors. 



## Hedonism, Usability and Beauty

In DN three separate scales were used, but parallel analysis suggests that these capture the same latent variable under both perspectives.

```{r}
parallel_analysis(D_HUB, 3, "P", "Hedonism, Usability and Beauty")

parallel_analysis(D_HUB, 3, "D", "Hedonism, Usability and Beauty")
```









# Discussion

Rating scales in Human Factors research are commonly used to discriminate between poor and good design options, rank designs, choose designs, or perform UX regression tests in continuous design cycles. Our logical argument is that the capability of a scale to rank designs can only be seen on multiple designs and using design-by-item response matrices. We called it the psychometric fallacy to use person-by-item response matrices in place. 

A simulation showed, that  the worst case can happen under the psychometric fallacy: excellent reliability is reported, when it actually is very poor. 


## Dev time implications




## Run time implications

When looking at real data from commonly used rating scales, a clear bias appears, but this time the Designometric perspective significantly improves reliability. This is good news for future users of these scales, as they can reduce their sample sizes.

Past users may have scratched their heads more often than we know. Techniques such as Cronbach $\alpha$ are commonly applied as quick sanity checks, whenever multi-item scales are used in run-time studies. Everyone in the past who used designometric scales and followed APA guidelines, was likely to the psychometric fallacy and experience some unnecessary headaches.

With better item-level reliability, experiments can also be streamlined by using a reduced set of items. Reducing a scale is basically item selection and several strategies are possible, as long as it is carried out on a design-by-item matrix.

A basic strategy is to use item-level reliability a ranking criterion. In many cases this may work well, but takes the risk of inadvertently reducing the precision in certain ranges of the scale. A well-constructed scale produces similarly precise measures in the lower, middle and high range on the whole scale. During development time this can be achieved by selecting items that are varying in strength, like "The interface is beautiful." is stronger than "... pleasant". By taking the average score


## Criticism of individual scales

One can rightfully argue that we used the rating scales outside their specification, as the encounters were brief and without interaction. This depends on the scale and on the purpose. Face processing is one of the fastest complex mechanisms in the human mind, which makes it legit to test it in quick succession on a screen. Similarly, the beauty judgement of a website is known to stabilize within 500ms, and the same can be expected for attractiveness. For more longitudinal feelings about designs, like hedonism, usability and credibility, the encounters were not valid. We will therefore not criticize these instruments, the point still standing that the two perspectives produce different results.

### Eeriness

### Beauty

### Attractiveness




## Future Applications 

A founding idea in usability engineering is that a good designers has learned to bridge the gap between the system model and the users mental model, cognitive skills and feelings. Emerging domains are often characterized by an innovation phase, where multiple design paths are explored, before this diversity converges into an oligarchy of mature sibling frameworks with established best practices and quality principles.   
Two emerging domains of human-technology interaction are currently gaining significant momentum: interactive deep learning models, such as chatbots and social robots. Both fields have in common that they tap much stronger into the social mind of users, which, next to being a cave of snakes, is mostly uncharted terrain. We can expect a wild growth of theoretical concepts and instruments to measure the respective latent variables.

[Simone]




## Towards Deep Designometrics

By comparing the two perspectives, we illustrated that designometric analysis can fully be done with standard psychometric tools, as long as one uses the correct response matrix. However, by reducing the designometric box to a flat matrix, we loose all information on users. Formally, it would even be possible to evaluate a designometric model on the responses of a *single user*, while the situation 

If the cube is collapsed to a  psychometric matrix, which can be used to estimate *user sensitivity*. Legitimate cases exist to use a designometric scale for psychometric purposes. For example, an instrument to measure trustworthiness of designs could be used to estimate faithfulness levels of participants in a study (or a training) on cyber security.

By flattening the designometric box one way, then the other, we still loose information that is needed to secure that items are truly well-behaved. In educational psychometrics *differential item functioning* is the idea that items must be fair and function the same for every tested person. This also is a  desirable property for a designometric scale, but a statistical model for verification would need individual parameters for participants, designs and items, simultaneously.

Schmettow(2021) proposed multi-level models for capturing designometric situations in their full dimension, which copuld be well-suited for run-time use or basic scale development. But, a full development workflow is lacking.

Another consideration is that the designometric encounter may not be end of story. For example, for comparing multi-purpose designs a researcher may want to add tasks as fourth population of interest. With the mentioned limitations, multi-level models extend to such a case (Schmettow, 2016, Egan's assumption).  For development-time purposes, Generalizability Theory may provide ... 

[Stephanie]

Another issue is to identify exploratory methods that can operate on designometric boxes. EFA is often used with CFA to find and then confirm candidate structures.

[Stephanie]


